---
title: "Fishash vignette"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Fishash vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We illustrate fishash on the [Schraivogel et al
2020](https://www.nature.com/articles/s41592-020-0837-5) CRISPRi
dataset which is also bundled with the [Python crispat
package](https://github.com/velten-group/crispat).

```{r load-packages, message = FALSE}
library(magrittr)
library(fishash)
library(ggplot2)
library(dplyr)
library(SummarizedExperiment)
library(Matrix)
```

Loading the data and printing out the first few entries, we can see it
is a sparse count matrix:

```{r load-data}
data(crispat_schraivogel)
crispat_schraivogel[1:10,1:10]
```

We pass this matrix to the `fishash()` function to call
presence/absence at 5% FDR:

```{r run-fishash}
res_fishash <- fishash(crispat_schraivogel, padj_cutoff = .05)
res_fishash
```

The return value is a SummarizedExperiment. The `assigned` assay is a
sparse boolean matrix indicating whether we call the guide present:

```{r show-assigned}
assay(res_fishash, 'assigned')[1:10, 1:10]
```

The colData shows the classification of each cell; the `demux_type`
column indicates whether the cell contained 0, 1, or 2+ assigned
guides, while the `assignment` column is a comma-delimited string
containing the guides called:

```{r show-coldata}
colData(res_fishash)
```

We can also inspect the log p-values from the Fisher test. Since we
use a one-sided test, this matrix is also sparse:

```{r show-logpval}
assay(res_fishash, 'log_pval')[1:10, 1:10]
```

Finally, the metadata shows the p-value cutoff selected by the FDR procedure:

```{r show-metadata}
metadata(res_fishash)
```

Below, we plot the nonzero log p-values together with the FDR cutoff,
coloring by the number of counts in the entry:

```{r plot-pvalues, fig.height=8, fig.width=8}
crispat_schraivogel %>%
  as('TsparseMatrix') %>%
  {data.frame(
    count=.@x,
    row_idx=.@i+1,
    col_idx=.@j+1
  )} %>%
  mutate(log_pval=assay(res_fishash, 'log_pval')[cbind(row_idx, col_idx)]) %>%
  mutate(log_pval_trunc=pmax(log_pval, -20)) %>%
  mutate(count=cut(count, c(0, 1, 2, 3, 5, 10, 30, 50, 100, Inf))) %>%
  ggplot(aes(x=-log_pval, fill=count)) +
  geom_histogram(bins=100) +
  geom_vline(xintercept=-metadata(res_fishash)$log_pval_cutoff, lty='dashed') +
  scale_x_continuous(trans='log1p', breaks=c(0,1,3,10,30,100,300,1000)) +
  scale_fill_viridis_d() +
  theme_classic(base_size=16) +
  theme(legend.position='bottom')
```

```{r session-info}
sessionInfo()
```
